Future challenges for Research Software Engineering

### 2025-08-11 future_challenges_for_RSEng.tex
Fix LaTeX
Proof-reading

Besides the direct impact in research software and further research output,
we see a number of risks.
The workshop participants suggested that the emergence of AI and automation might reduce critical thinking and imagination.
We address this in \authorRef{sec:complexity}.
Another risk is that the widespread use of AI tools will make community resources
such as \eg\ Stack Overflow and Wikipedia (which are used to train AI models) poorer, due to fewer incentives to maintain them.
Further risks that were later discussed include an increased replication crisis, a de-democratization of research, and the loss of trust in science.
Due to the non-deterministic nature of generative AI, we expect data generation and analysis to become less reproducible. Professional RSEs could safeguard reproducibility by designing, evaluating, and monitoring appropriate workflows.
Future research will require significant computing and storage resources, as well as useful data and models,
therefore it will be more driven by institutions able to provide these.
Finally, an increasing risk is the loss of trust in science,
in particular due to the automation of the peer-reviewing process~\cite{Naddaf2025}.
As more and more research will be written by researchers assisted by AI,
RSEs will spend more of their time in designing and reviewing software written by other humans and AI tools,
rather than repeatedly crafting similar individual elements.

>>>>

Besides the direct impact in research software and further research output,
we see a number of risks.
The workshop participants suggested that the emergence of AI and automation might reduce critical thinking and imagination.
We address this in \autoref{sec:complexity}.
Another risk is that the widespread use of AI tools will make community resources
such as \eg\ Stack Overflow and Wikipedia (which are used to train AI models) poorer, due to fewer incentives to maintain them.
Further risks that were later discussed include an intensified replication crisis, a de-democratization of research, and loss of trust in science.
Due to the non-deterministic nature of generative AI, we expect data generation and analysis to become less reproducible. Professional RSEs could safeguard reproducibility by designing, evaluating, and monitoring appropriate workflows.
Future research will require significant computing and storage resources, as well as useful data and models,
therefore it will be more driven by institutions able to provide these.
Finally, an increasing risk is the loss of trust in science,
in particular due to the automation of the peer-reviewing process~\cite{Naddaf2025}.
As more and more research will be written by researchers assisted by AI,
RSEs will be freed from repeatedly crafting similar individual elements, but will be spending more of their time in designing and reviewing software written by other humans and AI tools.
